How predictable is extinction? Forecasting species survival at million-year timescales

### 2019-06-10 doc/paper.tex
working through reviewer comments

\begin{abstract}
 A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton.) We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate) the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the near future. Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented. This has led to the suggestions that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}. The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.

Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how accurate or strong our predictions about the differences in extinction risk are.

Here we ask how precise risk predictions based on fossil data might be. Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions. However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g. ``future'', from the point of view of the model) extinction/survival events.

Extinction intensity (average rate) and selectivity (difference in risk between taxa) vary through time and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}. This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

@@

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an internally consistent taxonomic identification strategy -- as close to ideal data for this analysis as possible. The Neptune database records multiple phyla-scale taxonomic groups for over 60 million years, with incredible temporal resolution supported by the various age-models of the deep-sea cores from which the occurrences are recorded. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, foraminifera, and radiolarians.% -- these occurrences span the entire globe between 120 and 0 million years ago (Mya). 
Occurrences were filtered to include only those species with first occurrence was no more than 63 Mya. This filtering criterion excludes taxa that survived the K/Pg extinction or arose during its ecologically unusual recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

All fossil occurrences were assigned to 1 My bins based on the estimated age of the fossil occurrence as listed in the Neptune Database. After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the maximum great circle distance on an ellipsoid (i.e. the Earth) between any two occurrences of that species; this distance was measured in kilometers.

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My from 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

@@

See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other. A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}. Additionally, the full description of how these models were implemented and coded, including choice of priors, is available Section \ref{sec:model_est}.

\begin{table}[ht]
 \caption{Models and their definitions}
 \begin{threeparttable}
  {
   \def\arraystretch{1.5}
   \begin{tabular}{ l p{3cm} l l }
    Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
    \hline
    C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ age\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}})} \\
    V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ phylum) + (1 $|$ age/phylum)} \\ 
    CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff\tnote{g}\phantom{\textsuperscript{g}} +\\temp + temp\_lag\tnote{h}\phantom{\textsuperscript{h}} +\\(1 $|$ age/phylum)} \\
    VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff +\\temp + temp\_lag +\\(1 + range + range\_diff +\\temp + temp\_lag $|$ phylum) +\\(1 $|$ age/phylum)} \\
    \hline
   \end{tabular}
  }
  \begin{tablenotes}
  \item[a] See Equation \ref{eq:model} for full statistical model definition.
  \item[b] Species observation where 1 if time of last observation, otherwise 0.
  \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
  \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
  \item[e] Species are at observation in millions of years.
  \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
  \item[g] Change in geographic range since last observation.
  \item[h] Temperature at previous observation.
  \end{tablenotes}
 \end{threeparttable}
 \label{tab:model_def}
\end{table}

@@

In-sample forecasting means we are estimating how well our model predicts extinction probability for observations that that model was fit to. This is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.

@@

\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Finally, the model is fit to the first four folds combined and then is used to forecast the fifth fold. The results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

The relative adequacy of the four model variants was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used to measure the performance of classification models as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we interpret values between 0.7 and 0.8 could then be considered ``fair,'' and values between 0.6 and 0.7 as ``poor.''

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) AUC estimates for each of our four models. These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively. Models with a higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:auc_compare}
\end{figure}

@@

Depending on the taxon-model combination, there are between zero and 4 time intervals where our posterior distribution of in-sample AUC has a median value less than or equal to 0.5 (Fig. \ref{fig:auc_taxon_time}). However, this pattern is absent for the posterior estimates of in-sample AUC for Foraminfera and Radiolaria as fit by the VP model. In contrast, there are fewer periods of low model performance for calcareous nannoplankton and Dinoflagellates as estimated from our VP model than in those estimates from the other three model variations.

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
 \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:auc_taxon_time}
\end{figure}

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. The resulting distribution when all folds are combined is highly multimodal as expected given that they are fit to and estimated from different data sets with different numbers of observations \citep{ESL}. This multimodality increases with model complexity (Fig. \ref{fig:fold_auc}), most likely because the complex models allow for predictor effects to vary with time, allowing a greater range in possible parameter values which in turn yield a greater range of posterior predictions.

Comparison the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in posterior predictive distributions of AUC between the four model variations is reduced in out-of-sample prediction. For example, the VP model no longer has the greatest median AUC of the four models (Fig. \ref{fig:fold_auc}). For this reason the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, making interpretations based only on the median AUC estimates is incorrect -- our estimates are the full posterior, not individual points. Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig. \ref{fig:fold_auc}).

@@

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the Dinoflagellates, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:fold_auc_taxon_time}
\end{figure}

@@

We find that all of our models are expected to correctly forecasting the rank order of extinction probability for a randomly selected extinct-extant pair of future observations between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic. A extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

@@

We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result is not practically or scientifically significant. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance. Thus, including geographic range trajectory results in only very minor improvements in forecasting accuracy. This contrasts with findings from coral genera \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.

@@

The relative quality and consistency of our models' out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in our models is that we lack suitably detailed life history or ecological information for many marine micro- and nannoplankton. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.

>>>>

\begin{abstract}
  A tenet of conservation palaeobiology is that knowledge of past extinction patterns can help us to better predict future extinctions. Although the future is unobservable, we can test the strength of this proposition by asking how well models conditioned on past observations would have predicted subsequent extinction events at different points in the geological past. To answer this question, we analyze the well-sampled fossil record of Cenozoic planktonic microfossil taxa (foramanifera, radiolarians, diatoms, and calcareous nanoplankton). We examine how extinction probability varies over time as a function of species age, time of observation, current geographic range, change in geographic range, climate state, and change in climate state. Although the best-performing model includes time-varying effects and historical covariates (change in range and change in climate), the improvement in predictive power over models with constant effects and no historical covariates is very minor. Our models have a 70-80\% probability of correctly forecast the rank order of extinction risk for a random out-of-sample species pair, implying that determinants of extinction have varied only modestly through time. An important caveat is that human impacts may substantially disrupt range-risk dynamics so that the future will be less predictable than it has been in the past.

@@

The intensifying biodiversity crisis confronts conservation biologists with the difficult task of trying to predict which species are most threatened with extinction in the near future. Predicting which species will go extinct is difficult because reliable population and geographic range time series are typically known for only the past few decades in even the best-studied groups, and because few modern extinctions have been adequately documented. This has led to the suggestion that some risk assessments might be improved by incorporating palaeontological data \citep{Finnegan2015,Kiessling2016}. The fossil record preserves information about the full histories, include ultimate extinction, of thousands of lineages, and this information can help to augment the shorter-term higher-resolution data used to make risk assessments of extant taxa.

Past palaeobiological studies of extinction have frequently focused on identifying and measuring the effect of various potential predictors on extinction risk \citep{Harnik2011,Smits2015,Peters2008,Payne2007,Harnik2012,Ezard2011,Foote2006} or on how to identify or measure these effects \citep{Alroy2010a,Alroy2014,Alroy2001,Alroy2000,Alroy2000b,Foote2001}. This focus means that while we have a good understanding of which factors are strong and general determinants of extinction risk, we have less knowledge of how capable we are at predicting future extinction events.

We ask how accurate risk predictions based on fossil data are when estimating future extinctions. Because future extinctions are unobservable we cannot directly evaluate the ultimate performance of such predictions. However, we can address the question by taking a given point in the geological past, developing a predictive model based on extinction patterns prior to that point, and assessing the predictive performance of this model on unobserved (e.g. ``future'', from the point of view of the model) extinction/survival events.

Extinction intensity (average rate) and selectivity (difference in risk among taxa) vary through time, and the relative risk of extinction exhibited by different taxonomic groups and how that risk varies over time is an important dynamic which shapes the rate and structure of extinction \citep{Payne2007,Payne2016,Ezard2011,Smits2019}. This variation also raises a question: given that extinction intensity and selectivity change over time, how accurate are our assessments based on past events likely to be when applied to the future? Putting aside the important question of how human activities will alter the determinants of future extinction risk, we can address this uncertainty by specifically including and modeling the temporal variation in extinction risk across a range of extinction intensities and selectivities.

@@

For this exercise we chose to analyze one of the best-sampled and studied fossil records -- the Cenozoic record of skeletonized marine planktonic microorganisms (Foraminifera, Radiolaria, Diatoms, and Coccolithophores). These data are readily available through the Neptune database, an online repository of species occurrences obtained through the Deep Sea Drilling Program and the Ocean Drilling Project \citep{Lazarus1994,SpencerCervato1999}. This database provides abundant samples in space and time, a high degree of temporal resolution for the entirety of the Cenozoic, and has an taxonomy synonymization strategy for delaing wiht 50+ years of taxonomic opinion \citep{Lazarus1994} -- as close to ideal data for this analysis as possible. Analyzing patterns of extinction and global occurrence at fine temporal scales means we can better elucidate how well we can predict species extinction at human-relevant scales.

@@

We analyzed microfossil occurrence information from the Neptune Database \url{http://www.nsb-mfn-berlin.de/nannotax} \citep{Lazarus1994,SpencerCervato1999}. This occurrence-based dataset includes calcareous nannoplankton, diatoms, planktonic foraminifera, and radiolarians. Occurrences were filtered to include only those species who's first occurrence is no earlier than 63 Mya (millions of years ago). This filtering criterion excludes taxa that survived the K/Pg extinction or arose during this recovery interval, and ensures that our occurrence histories fully overlap with the temperature time-series used as a potential extinction risk predictor (see below).

All fossil occurrences were assigned to 1 My (million year) bins based on the estimated age of the fossil occurrence as listed in the Neptune Database. After binning, each taxon's geographic range was calculated for each of the 1 My bins in which it occurred. Geographic range was calculated as the minimum spanning tree distance between all observations of that taxon during that temporal bin; this distance was measured in kilometers. Minimum spanning tree distance was caluclated using the \texttt{GeoRange} package for R \citep{GeoRange}.

Average global temperature of each 1 My bin was calculated from estimates based on Magnesium/Calcium isotope ratios \citet{Cramer2011}. We use Mg/Ca rather than oxygen isotopes to avoid confounding effect of varying ice-volume -- this property is of particular importance for this analysis as polar ice-caps develop midway through the Cenozoic. Estimating temperature over long periods of time from Mg/Ca ratios also suffers from complications because Mg/Ca based temperature estimates requires benthic foram and seawater Mg/Ca isotope ratio information. Because seawater Mg/Ca ratio has changed over time, the method to estimate temperature used \citet{Cramer2011} attempts to account for this unknown in order to obtain accurate, albeit uncertain, temperature estimates. Our data source, \citet{Cramer2011}, estimated temperature for every 0.1 My interval between 0 to 63 Mya. The temperature estimate for each 1 My interval was calculated as the mean of all estimates within that interval.

These temperature estimates are measured from benthic forams, and is an estimate of deep water temperature. The organisms included in this study are all planktonic, so these temperature estimates do not necessarily represent the actual environment these organisms inhabit. However, detailed ocean surface temperature estimates do not exist for the entirety of the Cenozoic. So, while deepwater temperature estimates are not ideal for understanding the extinction risk of planktonic taxa, we interpret our estimate of temperature as reflecting global climate state as opposed to a taxon-specific ecological descriptor.

@@

See Table \ref{tab:model_def} for further explanation of how the four models we considered differ from each other. A complete description of the statistical model used in this analysis is available in Section \ref{sec:model_desc}. Additionally, the full description of how these models were implemented and coded, including choice of priors, is available in Section \ref{sec:model_est}.

\begin{table}[ht]
 \caption{Models and their definitions}
 \begin{threeparttable}
  {
   \def\arraystretch{1.5}
   \begin{tabular}{ l p{3cm} l l }
    Code & Description & Covariates & R Formula Syntax\tnote{a}\phantom{\textsuperscript{a}} \\
    \hline
    C & Constant effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event\tnote{b}\phantom{\textsuperscript{b}} $\sim$ range\tnote{c}\phantom{\textsuperscript{c}} + temp\tnote{d}\phantom{\textsuperscript{d}} +\\(1 $|$ time\tnote{e}\phantom{\textsuperscript{e}}/phylum\tnote{f}\phantom{\textsuperscript{f}}) +\\(1 $|$ age\tnote{g}\phantom{\textsuperscript{g}}/phylum)} \\
    V & Varying effects, no historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + temp +\\(1 + range + temp $|$ time/phylum) + (1 $|$ age/phylum)} \\ 
    CP & Constant effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ + range\_diff1\tnote{g}\phantom{\textsuperscript{h}} + range\_diff2\tnote{h}\phantom{\textsuperscript{h}} + range\_diff3\tnote{h}\phantom{\textsuperscript{h}} + \\temp + temp\_lag\tnote{i}\phantom{\textsuperscript{i}} +\\(1 $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    VP & Varying effects, historical cov. & \parbox[t]{0.25\textwidth}{Geographic range,\\change in geographic range, temperature,\\previous temperature} & \parbox[t]{0.33\textwidth}{event $\sim$ range + range\_diff1 + range\_diff2 + range\_diff3 + \\temp + temp\_lag +\\(1 + range + range\_diff1 + range\_diff2 + range\_diff3 +\\temp + temp\_lag $|$ time/phylum) +\\(1 $|$ age/phylum)} \\
    \hline
   \end{tabular}
  }
  \begin{tablenotes}
  \item[a] See Equation \ref{eq:model} for full statistical model definition.
  \item[b] Species observation where 1 if time of last observation, otherwise 0.
  \item[c] Species geographic range in log km\(^2\). Mean centered, scaled to sd = 1.
  \item[d] Global temperature in degrees C. Mean centered, scaled to sd = 1.
  \item[e] Time of observation.
  \item[f] Taxonomic group of species (i.e. Foraminifera, Diatoms, Radiolarians, Calcareous nannoplankton).
  \item[g] Age at observation.
  \item[h] Change in geographic range since last observation (number indicates how lags).
  \item[i] Temperature at previous observation.
  \end{tablenotes}
 \end{threeparttable}
 \label{tab:model_def}
\end{table}

@@

In-sample forecasting is a posterior predictive check in that we are comparing the posterior predictive distribution to our observed data. This is a type of sensitivity analysis for understanding the quality of model fit to the data, and how well our model predicts extinction probability for observations that we used to fit out model initially. In-sample forecasting measures, however, are not necessarily good estimates of the model's ability to predict data from the future \citep{ESL}.

@@

\textit{k}-fold cross-validation for time series follows a specific sequence of procedures \citep{Arlot2010,Bergmeir2016}. First, the model is fit to the first fold (time segment), and the posterior estimates of that fit are then used to forecast the extinction probability of the second fold (i.e. the future). Then the model is fit to the combined first and second folds, and the posterior estimates of that fit are used to forecast the extinction probability of the third fold. Continuing, the model is then fit to the first three folds combined and is then used to forecast extinction probabilities for the fourth fold. Next, the model is fit to the first four folds combined and then is used to forecast the fifth fold. This process continues until $k - 1$ folds are included in the fitting the model and the final fold is predicted from this model. When combined, the results from these forecasts are then combined to yield our estimate of expected out-of-sample performance.

Some readers may be familiar with metrics such as AIC, WAIC, or LOOIC for comparing model fit. These metrics are approximations of out-of-sample predictive error as estimated by cross-validation \citep{ESL}. Because we've estimated out-of-sample performance using cross-validation, these measures would be redundant, and in the case on single value measures such as AIC, much less informative than an entire posterior distributions. When out-of-sample estimates from a complex model are noticeably worse than out-of-sample estimates from a simpler model, than this is evidence for the complex model being overfit to the data \citep{ESL}. In contrast a high degree of similarity between out-of-sample and in-sample estimates indicates that the model is not overfit to the data (though it is not necessarily an adequate descriptor of the data).

The relative adequacy of the four model variants was compared using the area under the receiver operating characteristic curve or AUC \citep{Fawcett2006,Mason2002}. This measure is commonly used to measure the performance of classification models as it has the desirable characteristic of comparing the model's true positive rate with its false positive rate, as opposed to accuracy which only considers true positives. AUC ranges between 0.5 and 1, with 0.5 indicating no difference in classification from random and 1 indicating perfect classification. AUC can be interpreted as the probability that our model correctly ranks the relative extinction risks of a randomly selected extinct-extant species pair \citep{Fawcett2006,Mason2002}. AUC values of approximately 0.8 or greater can be considered ``good'' \citep{ACCDA}, so we consider values between between 0.7 and 0.8 as ``fair,'' and values between 0.6 and 0.7 as ``poor.''

@@

% ROC model comparison
\begin{figure}[ht]
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_hist_full}
  \caption{In-sample}
  \label{fig:auc_hist}
 \end{subfigure}
 \begin{subfigure}[ht]{0.45\textwidth}
  \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_full}
  \caption{Out-of-sample}
  \label{fig:fold_auc}
 \end{subfigure}
 \caption{In-sample (\ref{fig:auc_hist}) and out-of-sample (\ref{fig:fold_auc}) area under the receiver operating charactered curve (AUC) estimates for each of our four models. These estimates are calculated from the models posterior predictive distribution or from predictions made to new data, respectively. Marked below the posterior distributions are the median AUC and 50\% and 80\% posterior intervals for all observations in our dataset. Models with higher AUC values indicate better performance over models with lower AUC values. AUC is bounded between 0.5 and 1. See Table \ref{tab:model_def} for an explanation of the four models (C, V, CP, VP).}
 \label{fig:auc_compare}
\end{figure}

@@

The in-sample forecasts from our four models over time are virtually identical within taxa (Fig. \ref{fig:auc_taxon_time}). Our in-sample forecasts for Diatoms are the weakest of the taxonomic groups as all four models have an equal number of intervals with no predictive power (AUC approximately 0.5). In contrast, our best in-sample forecast results are for Radiolarians where there is at most 1 interval with almost no predictive power. Ultimately, our in-sample forecasts are virtually identical across models.

\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/auc_taxon_time_full}
 \caption{Comparison of in-sample forecasting performance measured by AUC for each of the four models, arranged over time and by taxonomic group. These estimates reflect each model's fit to the various taxonomic groups over time. The black line corresponds to the median AUC value, while the envelopes correspond to multiple credible intervals as indicated in the legend. In all cases, higher AUC values indicate greater predictive performance versus lower AUC values. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:auc_taxon_time}
\end{figure}

@@

Expected out-of-sample forecasting performance was estimated using five-fold cross-validation for time series \citep{Arlot2010,Bergmeir2016}. Comparing the in-sample AUC estimates to the expected out-of-sample AUC estimates reveals a similar range in performance for all models (Fig. \ref{fig:auc_hist}, \ref{fig:fold_auc}). Interestingly, the differences in posterior predictive distributions of AUC between the four model variations is reduced in out-of-sample prediction. For example, the VP model no longer has the greatest median AUC of the four models (Fig. \ref{fig:fold_auc}). For this reason the rank order of median out-of-sample AUC is different from the rank order of median in-sample AUC. However, making interpretations based only on the median AUC estimates is incorrect -- our estimates are the full posterior, not individual points. Because of this, the four models are effectively indistinguishable in their out-of-sample forecast performance (Fig. \ref{fig:fold_auc}).

@@

In the analysis of the in-sample forecast performance of the four models, we noted that there were time intervals where our predictions were no better than random (Fig. \ref{fig:auc_taxon_time}). This occurrence is generally much rarer for the posterior distribution of AUC from the out-of-sample forecasts. The major exception to this pattern are our estimates for the diatoms, which have at least one time interval for all four models in which the median AUC of the out-of-sample forecasts were no better random. The only other group for which median posterior predictive estimate of out-of-sample AUC reaches 0.5 is calcareous nannoplankton, and then only with the V model.
\begin{figure}[ht]
 \centering
 \includegraphics[width=\textwidth,height=0.5\textheight,keepaspectratio=true]{../results/figure/fold_auc_taxon_time_full}
 \caption{Comparison of out-of-sample AUC values over time as aggregated by taxonomic group for each of the four models. The AUC of the individual My intervals within each fold is plotted to highlight the heterogeneity in performance within and between folds. This presentation decomposes each of the 12-million year folds by each of the taxonomic groups into the predictions made for each of the million-year intervals. The black line corresponds to the median AUC estimate, with the envelopes corresponding to multiple credible intervals as indicated in the legend. The grey intervals mark the geologic ages of the Cenozoic. See Table \ref{tab:model_def} for a description of each of the four models.}
 \label{fig:fold_auc_taxon_time}
\end{figure}

@@

We find that all of our models are expected to correctly forecast which species of randomly selected extinct-extant pair is more likely to go extinct between 70\% to 80\% of the time (Fig. \ref{fig:fold_auc}). One of the most striking aspects of these results is the similarity in forecasting performance for in-sample and out-of-sample observations. A slight decrease in performance when dealing with out-of-sample observations makes sense: each of the models fit during cross-validation is based on fewer data than the model fit on the full data (between 1/5th to 4/5ths of the original). Additionally, a potential decrease in precision when forecasting the future of extinction risk is to be expected as the future will always differ from the past in some respect. However, the similarly between the in-sample and out-of-sample results indicates that our model is fairly robust to variation in extinction intensity over the Cenozoic. An extremely important caveat, of course, is that human impacts may substantially alter present and future extinction risk dynamics relative to the average Cenozoic state, so that the future may become less predictable than it has been in the past \citep{Harnik2012a,Finnegan2015}.

@@

We also find that our four models are practically identical in their ability to make in-sample and out-of-sample forecasts. Although the in-sample AUC estimates differ between models, all of these estimates are in a narrow range of possible AUC values (Fig. \ref{fig:auc_hist}). While the ``best'' model does include the historical covariates and allows all covariate effects to vary over time, its practical difference in performance versus the other models is negligible (Fig. \ref{fig:fold_auc}). Thus even though the VP model has a statistically greater AUC for in-sample forecasts than the other three models, this result has no practical significance because the predictions are functionally identical. The results of the out-of-sample forecasts illustrate this point in full, as all four models have functionally identical out-of-sample performance. Thus, including geographic range trajectory results in only very minor improvements in forecasting accuracy. This contrasts with previous findings using geological stages as their time unit \citep{Kiessling2016} and suggests that further investigation of other taxa, timescales, and environments may help to elucidate the conditions under which past geographic range trajectories are most informative about future extinction risk.

@@

The relative quality and consistency of our models' out-of-sample forecasting performance is encouraging given that these estimates are based on very limited biological and environmental information about the studied taxa. Even our most complex models only account for a few simple aspects of geographic range, prior history, and phylogenetic affinity. The principal reason we were not able to include more biological information in the models used here is because we lack additional life history or ecological information for many of the marine micro- and nannoplankton included in this study. Foraminifera are an exception to this problem as aspects of life history, ecology, and physiology are known for many foram species \citep{Ezard2011}. However, comparable information does not exist for all foram species, nor does this type of data exist for the other three taxonomic groups studied here. Future analyses including this type of information and focused more narrowly on the foraminifera may be informative.

In summary, our results suggest that models trained on prior extinction/survival patterns do modestly well at predicting relative extinction probability of randomly selected species pairs based on a small number of taxonomic, geographic, and historical predictors. While change to in geographic range and past temperature information is not found to greatly improve our ability to estimate future extinction risk, we do find that the immediate past is a modest predict of the immediate future. These results are comparable to conservation determinations because both are expressed in terms of an relative risk. The results of this simple exercise suggest that conservation decisions would be bolstered by including fossil data.
